\large{Vipassana for Hackers} \\ \Huge{Paper Three: Why Meditate?} \\ \large\textit{Version 0.1}

### 2019-11-25 paper-three/paper-three.tex
meditation planes for paper 3

\begin{abstract}
 \todo{Meditation requires a time commitment. Like other activities we consider
   worthy of our time for their benefits to our health, such as sufficient sleep, exercise, fresh air, and a
   healthy diet, it is often the (rational) first step of individuals considering a
   meditation practice to ask: Why should I bother? What are the outcomes of
   meditation? And do the benefits of those outcomes outweigh the time meditation
   costs the practitioner? This paper answers those questions as they pertain to
   meditation in general, and to Vipassana, specifically.}
\end{abstract}

@@

\section{The Mundane and The Supramundane}

There are two fields of human experience and the following analysis of the value
provided by different meditation techniques will break down all points into these two
categories. \textit{The Mudane}
in this context refers not to the tedious but to the earthly, the material. Most people will begin
meditating for reasons in the mundane field simply because most people have never
experienced the supramundane field. \textit{The Supramundane} in this context refers not to
spirituality or religion but experiences which transcend the material, physical world.

@@

@@

Meditation objectively improves sleep. Black et al. found in a 2015 study that this
improvement is is not simply over the baseline, however, but also an improvement
above and beyond what can be achieved through Sleep
Hygiene Education (SHE). \cite{black2015} The difficulty with this statement, of
course, is defining the term ``meditation''. In this particular study, the meditation
in question is \textit{Mindfulness}, as taught in the UCLA Mindfulness
Course. \cite{uclamaps} Mindfulness is a very accessible form of meditation, varying
in format. It usually involves multiple techniques, including \textit{Open
  Monitoring} which is characterized by
openness to whatever is happening in the present moment in a variety of postures
(sitting, standing, walking, eating, etc.), \textit{Focused Attention} which is
usually a narrow breath awareness meditation, and \textit{Loving-Kindness} in which
meditators actively direct compassion to themselvse and others. All three of these
practices mirror the three meditation practices taught in Vipassana. However, in Vipassana, Open
Monitoring finds its parallel in objective observation of sensation internal to the
body, as we will see later.

@@

\begin{itemize}
  \item reset frame of reference outside oneself, outside one's own lifetime: ``trees
    for god'' and obvious karma (sidu/booga smoking)
  \item Time: Nat Friedman's blog post?
  \item clarification: ``isn't that what makes us human?'' (emotions) --- rather,
    what makes us animal
  \item ``How to Change Your Mind'' --- ego dissolution --- specific advantages?
\end{itemize}

@@

>>>>

\begin{abstract}
 Meditation requires a time commitment. Like other activities we consider
 worthy of our time for their benefits to our health, such as sufficient sleep, exercise, fresh air, and a
 healthy diet, it is often the (rational) first step of individuals considering a
 meditation practice to ask: Why should I bother? What are the outcomes of
 meditation? And do the benefits of those outcomes outweigh the time meditation
 costs the practitioner? This paper answers those questions as they pertain to
 meditation in general, and to Vipassana, specifically.
\end{abstract}

@@

\section{The Mundane vs. The Supramundane}

There are two fields of human experience and the following analysis of the value
provided by different meditation techniques will break down all points into these two
categories. \textit{The Mundane}
in this context refers not to the tedious but to the earthly, the material. Most people will begin
meditating for reasons in the mundane field simply because most people have never
experienced the supramundane field. \textit{The Supramundane} in this context refers not to
spirituality or religion but experiences which transcend the material, physical
world. Put another way, mundane experiences are those which can be
described. Supramundane experiences, due to their very nature of transcending the
material world, are ineffable. Just because an experience is ineffable, however, does
not mean conversation cannot exist around it. An experience of dimethyltryptamine
(DMT) may be ineffable but it is almost a guarantee that someone using this drug will
talk about it afterward. In this way, we will discuss the consequences of
supramundane experiences (and their value) toward the end of this paper.

@@

In \textit{Paper One: Curious Mechanics}, two sample spectrums of meditations were
listed. One included activities which were simply meditative, rather than anything
which could be considered a formal meditation practice. These included sports,
playing a musical instrument, or yogic practices such as asanas and pranayams. We
will ignore such activities here. The following plane attempts to roughly locate
meditation practices between focus and insight on two axes, rather than one.

\begin{figure}[H]
  \centering
  \includegraphics[width=5cm]{images/focus-insight-plane.png}
  \caption{Meditation Quadrants}
  \label{fig:meditation-quadrants}
\end{figure}

The locations of different meditation practices on the graph are only for
illustrative purposes. There are many other axes available such as a practitioner's
experience level, intensity of practice, and an individual's physiology and mental
health. However, this plane helps ground a comparison of different meditation
techniques for the rest of the paper. The three mindfulness meditation categories are
Focused Attention Meditation (FAM), Open Monitoring Meditation (OMM), and
Loving-Kindness Meditation (LKM). These will be explained as mindfulness meditations
come up in discussion later.

Returning to the topic of the supramundane, the following spectrum illustrates where
each of these practices approximately falls, across that dichotomy.

\begin{figure}[H]
  \centering
  \includegraphics[width=5cm]{images/mundane-supramundane-plane.png}
  \caption{Mundane vs. Supramundane Meditations}
  \label{fig:mundane-vs-supramundane}
\end{figure}

Meditations within the mundane sphere are easier to understand and easier to
teach. As a consequence, they are easier to submit to rigorous scientific
study. Where the broad categories of FAM, OMM, and LKM have been studied
scientifically, we will observer respective parallels in Anapana, Vipassana, and
Metta meditations.

@@

Meditation objectively improves sleep. Black et al. found in a 2015 study that this
improvement is is not simply over the baseline, however, but also an improvement
above and beyond what can be achieved through Sleep
Hygiene Education (SHE). \cite{black2015} The difficulty with this statement, of
course, is defining the term ``meditation''. In this particular study, the meditation
in question is \textit{Mindfulness}, as taught in the UCLA Mindfulness
Course. \cite{uclamaps} Mindfulness is a very accessible form of meditation, varying
in format according to the instructor. It usually involves multiple techniques,
including \textit{Open Monitoring}, which is characterized by
openness to whatever is happening in the present moment in a variety of postures
(sitting, standing, walking, eating, etc.); \textit{Focused Attention}, which is
usually a narrow breath awareness meditation; and \textit{Loving-Kindness}, in which
meditators actively direct compassion to themselvse and others. All three of these
practices mirror the three meditation practices taught in Vipassana. However, in Vipassana, Open
Monitoring finds its parallel in objective observation of sensation internal to the
body, as we will see later.

@@

@@

\section{The Supramundane Sphere of Experience}

\begin{itemize}
  \item reset frame of reference outside oneself, outside one's own lifetime: ``trees
    for god'' and obvious karma (sidu/booga smoking)
  \item Time: Nat Friedman's blog post?
  \item bible; ``a perfect set of rules''; nature itself <=> perception; ``perfect
    perception (vision)''; inner perception (sensation mirror) --- harris v. peterson
  \item clarification: ``isn't that what makes us human?'' (emotions) --- rather,
    what makes us animal
  \item ``How to Change Your Mind'' --- ego dissolution --- specific advantages?
  \item seeing oneself as aggregate and the path from singular => aggregate
    (neuroscience behind hemispheres)
\end{itemize}
